Live Functional Programming with Typed Holes

### 2018-03-28 abstract.tex
draft of sec 3.0 through 3.1

@@

\emph{Live programming environments} aim to provide programmers (and, in some
cases, audiences) with continuous feedback about a program's dynamic behavior
throughout the editing process.
%
Unfortunately, because programming languages typically assign dynamic meaning
only to programs that are \emph{complete}---syntactically well-formed and free
of static type errors---live programming features are often severely restricted,
if available at all, when programs are incomplete.
%
Incomplete edit states are frequently encountered when writing programs, so they
are, arguably, when live programming services would be most valuable.
%
Defining a semantics for incomplete functional programs would help bridge the
gap between well-formed states and allow live programming environments to
provide continuous, uninterrupted, and meaningful feedback.

Recent work by \citet{popl-paper} makes progress toward this goal by developing
a type system for lambda terms with \emph{holes} in both expression and type
position, with empty holes standing for missing expressions or types, and
non-empty holes operating as ``membranes'' around type inconsistencies.
%
The contribution of this paper is a corresponding \emph{dynamic semantics for
functional programs with holes}.
%
Rather than aborting with an exception when evaluation encounters a hole (as in
several existing systems), evaluation proceeds ``around'' expression holes,
performing as much computation as is possible and tracking the dynamic
environment around each hole as it flows through the program.  These accumulated
hole environments can be reported directly to the programmer by the editor as
they work to fill holes.  They also enable an ``edit-and-resume'' feature that
avoids the need to restart evaluation after edits that amount to hole filling.

Formally, our semantics combines features from gradual type theory (to handle
incomplete types) and contextual modal type theory (CMTT, which provides a
logical foundation for hole environment tracking) with novel structure necessary
to prove key metatheoretic properties, including (1) a type safety property that
accounts for expression holes (i.e. free metavariables, from the perspective of
CMTT), and (2) commutativity and confluence properties that guarantee that the
``edit-and-resume'' feature is sound.
%
We have mechanized our formal development using the Agda proof assistant.
%
We additionally describe a simple reference implementation, called Hazelnut
Live, that exposes a language of structured edit actions that insert holes as
necessary to guarantee that every edit state has some (possibly incomplete)
type, based on the edit action calculus in the prior work by \citet{popl-paper}.
%
Taken together, the result is a proof-of-concept live functional programming
environment where dynamic feedback is truly continuous, i.e. it is available for
every possible edit state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

>>>>

% !TEX root = hazelnut-dynamics.tex

@@

\emph{Live programming environments} aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of static type errors, whereas 
%
during much of the editing process, the program is not yet complete. As a result,  
live programming services flicker in and out, 
in many cases going out for substantial lengths of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.
This paper confronts this problem from type-theoretic first principles by developing a 
\emph{dynamic semantics for incomplete functional programs}, 
based in part on the static semantics for incomplete functional programs
developed in recent work by \citet{popl-paper}. 
%
Incomplete functional programs are modeled as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as ``membranes'' around static and dynamic type inconsistencies. 
%
Rather than aborting with an exception when evaluation encounters any of these holes (as in
several existing systems), evaluation proceeds ``around'' the holes,
performing as much of the remaining computation as is possible and tracking the 
closure around each hole instance as it flows through the program.  These hole closures
 can be reported both directly and indirectly, via various editor services, to help the programmer decide how to fill a hole. 
They also enable an ``edit-and-resume'' feature that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics combines features from gradual type theory (to handle
type holes) and contextual modal type theory (which provides a
logical foundation for hole closures) with novel formal structure necessary
to continue evaluation past holes and to prove key metatheoretic properties. 
% , including 
% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
% ``edit-and-resume'' feature is sound.
%
We have mechanized the core formal development using the Agda proof assistant.
%
We additionally describe a simple reference implementation, called \HazelnutLive{},
that inserts holes as
necessary during the editing process to guarantee that every edit state has some (possibly incomplete)
type, based in part on the edit action calculus in prior work by \citet{popl-paper}.
%
Taken together with the type safety property that this paper establishes, the result is 
a proof-of-concept live typed functional programming environment where dynamic feedback 
is truly continuous, i.e. it is available for every possible edit state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

### 2018-03-28 calculus.tex
draft of sec 3.0 through 3.1

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\cy{exposition}

\subsection{Calculus}

\subsubsection{Syntax}

@@

Shorthand:

$$
\dcastthree{\dexp}{\htau_1}{\htau_2}{\htau_3} \defeq
  \dcasttwo{\dcasttwo{\dexp}{\htau_1}{\htau_2}}{\htau_2}{\htau_3}
$$

\rkc{this syntactic sugar is used in four places: ITCastSucceed, ITCastFail,
ITGround, and ITExpand. that's not many, and those rules don't look much more
cluttered without the sugar, so consider eliminating it. if so, just toggle the
definition of the dcastthree macro to the unsugared option.}

Explain that $c$ is a placeholder constant (i.e., unit), $b$ is a placeholder
type (i.e., unit).

\subsubsection{Statics}

\input{fig-type-consistency}
\input{fig-bidirectional-typing}

{\color{gray}\blindtext}
{\color{gray}\blindtext}
{\color{gray}\blindtext}

A typing context, $\hGamma$, is a list of bindings $x : \htau$.

\subsubsection{Expansion}

@@

{\color{gray}\blindtext}
{\color{gray}\blindtext}
{\color{gray}\blindtext}

@@

\subsubsection{Dynamics}

\subsubsection{Examples}

\clearpage
%% \subsection{Metatheory}
\subsection{Metatheory: Type Soundness}

@@

@@

\input{fig-complete}

@@

>>>>

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}. Sec.~\ref{sec:agda-mechanization} provides some additional details on this mechanization and the supplemental material includes the full mechanization. % The proofs of the metatheorems are established by this mechanization; the proofs are only very briefly outlined in this section.

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes, which we refer to as H-types and H-expressions, respectively. We distinguish between \emph{external} H-expressions, $e$, and \emph{internal} H-expressions, $d$. External H-expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for examples of manual, semi-automated and fully automated hole entry methods). Each well-typed external H-expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal H-expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}). We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. This machinery is initialized by the expansion step, rather than entered explicitly by the programmer. In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.

@@

% \rkc{this syntactic sugar is used in four places: ITCastSucceed, ITCastFail,
% ITGround, and ITExpand. that's not many, and those rules don't look much more
% cluttered without the sugar, so consider eliminating it. if so, just toggle the
% definition of the dcastthree macro to the unsugared option.}

\subsection{Static Semantics of the External Language}
\label{sec:external-statics}

\input{fig-bidirectional-typing}
\input{fig-type-consistency}

Let us start with a brief overview of the static semantics of the \HazelnutLive external language. H-types, $\htau$, classify external H-expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up). The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.

The benefit of specifying the \HazelnutLive external language bidirectionally, rather than as a type assignment system, is that the programmer does not need to annotate each hole with a type. Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole names are new to \HazelnutLive). Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$. If an empty hole appears where an expression of some other type is expected, e.g. under an explicit ascription $e : \htau$ (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$. Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}. Two types are consistent if they differ only up to type holes in corresponding positions. Note that type consistency is reflexive and symmetric but it is not transitive (quite unlike subtyping, which is anti-symmetric and transitive, though subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}). The hole type is consistent with every type, and so, by the subsumption rule, empty expression holes can appear where an expression of any type is expected.

Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly. Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type. Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}

For the familiar forms of the lambda calculus, the rules again follow the prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.

Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.  Half-annotated lambdas can appear in synthetic position (where an expected type is not known, e.g. at the top-level) according to Rule {SLam}, which is standard \cite{Chlipala:2005da}. Unannotated lambdas can only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were the type $\tarr{\tehole}{\tehole}$. To avoid the need for two separate rules, the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in Fig.~\ref{fig:arrmatch} produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}. This allows us to specify a single rule, Rule {ALam}, governing unannotated lambdas. Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might also define a synthetic rule governing unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.

The matched arrow type judgement also appears in the rule governing function application, Rule {SAp}, which treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$.

\subsection{Expansion}
\label{sec:expansion}

@@

@@

@@

\subsection{Dynamic Semantics}
\label{sec:evaluation}

@@

@@

\subsection{Agda Mechanization}
\label{sec:agda-mechanization}
